Multifunctional Lipid Probes Background

### 2025-01-30 LipidProbe/EnrichedHitsComparison.qmd
pre-meeting update

<!-- Data wrangling, will comment out most of this when it's all done. 
Pulling the enriched hits from each probe, then making a long dataframe that can be plopped into a Venn diagram adjacent visualization -->
```{r setup, echo=FALSE,include=FALSE}
library(dplyr)
library(tidyr)
library(readr)

# # Making a big shared dataset with all of the TMT-quantified lipid probes - this is a little hand-wavey because some were run at different times... TBD if we want to keep this.
# PA_data <- read_csv("LipidProbe/DataSets/PA_HeLa_AT_2025.csv") |>
# 	select(LipidProbe, gene_name, logFC, pvalue, hit_annotation)

@@

:::{.border}
:::{#fig-ShinyEmbed}
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Shiny App Embedding</title>
  <style>
    iframe {
      width: 100%;
      height: 90vh;
      border: none;
    }
  </style>
</head>
<body>
  <iframe src="https://gaelen-guzman.shinyapps.io/lipidinteractomics_shinyapp/"></iframe>
</body>
</html>
```

Black proteins are unenriched in respect to both probes; Purple proteins are "enriched" with respect to only the probe on the x-axis; Green proteins are "enriched" with respect to only the probe on the y-axis; and Orange proteins are "enriched" in both probes. Depicted are only the proteins present in the datasets of both probes.
:::
:::

>>>>

## Ratiometric data

For a select subset of the lipid probes depicted on this repository, one can directly compare their enrichment of proteins. In the case of the trifunctionalized PE, PA, Sph, Spa, 8-3 FA, and 1-10 FA probes, all were analyzed using quantitative mass spectrometry via [Tandem Mass Tagging](@sec-TMT).

Use the dropdown menu below to select two lipid probes to plot against each other and visualize the enriched hits of each probe. Black proteins are unenriched in respect to both probes; Purple proteins are "enriched" with respect to only the probe on the x-axis; Green proteins are "enriched" with respect to only the probe on the y-axis; and Orange proteins are "enriched" in both probes. Depicted are only the proteins present in the datasets of both probes.

<!-- Data wrangling, will comment out most of this when it's all done. 
Pulling the enriched hits from each probe, then making a long dataframe that can be plopped into a Venn diagram adjacent visualization -->
```{r setup, echo=FALSE,include=FALSE}
library(dplyr)
library(tidyr)
library(readr)
library(ggvenn)
library(crosstalk)
library(VennDiagram)
library(htmlwidgets)
library(plotly)
library(DT)
# # Making a big shared dataset with all of the TMT-quantified lipid probes - this is a little hand-wavey because some were run at different times... TBD if we want to keep this.
# PA_data <- read_csv("LipidProbe/DataSets/PA_HeLa_AT_2025.csv") |>
# 	select(LipidProbe, gene_name, logFC, pvalue, hit_annotation)

@@

:::{.border}
```{=html}
<!DOCTYPE html>
<html lang="en">
<head>
  <meta charset="UTF-8">
  <meta name="viewport" content="width=device-width, initial-scale=1.0">
  <title>Shiny App Embedding</title>
  <style>
    iframe {
      width: 100%;
      height: 90vh;
      border: none;
    }
  </style>
</head>
<body>
  <iframe src="https://gaelen-guzman.shinyapps.io/lipidinteractomics_shinyapp/"></iframe>
</body>
</html>
```
:::

## Shared hits

```{r, echo=FALSE, warning=FALSE, include=FALSE}
# Making a dataset for a venn diagram maker! This will all be commented out at the end.

# PA_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/PA_HeLa_AT_2025.csv") |>
#   filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# PE_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/PE_HeLa_AT_2025.csv") |>
# 	filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# Spa_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/Spa_Huh7_SF_2024.csv") |>
# 	filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# Sph_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/Sph_Huh7_SF_2024.csv") |>
# 	filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# FA_8_3_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/8-3_FA_Huh7_SF_2024.csv") |>
# 	filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# FA_1_10_data <- read_csv("/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/1-10_FA_Huh7_SF_2024.csv") |>
# 	filter(hit_annotation == "enriched hit") |>
# 	select(LipidProbe, gene_name)

# write.csv(combinedDataset, "/Users/gaelenguzman/LipidInteractomics_Website_local/LipidInteractomics_Website/LipidProbe/DataSets/combinedProbeDatasets_VennDiagram.csv")
```

### 2025-01-30 TrifunctionalLipidProbesOverview.qmd
pre-meeting update

---
title: "Multifunctional Lipid Probes Background"
author: "Gaelen Guzman"
format: 
  html:
    embed-resources: false
standalone: true
crossref: 
  ref-hyperlink: true # (default is true)
---

@@

## Introduction: Why study lipid interactomes?

* Because at its core biology is just complex chemistry, all biological reactions are mediated through the physical interaction of two or more molecules

* Lipid::Lipid interactomes can provide insight into which of these physical contacts mediate these membrane properties. Lipids are most commonly enmeshed within membrane bilayers and lipid-lipid interactions confer membrane properties such as rigidity or detergent resistance.

* In recent decades, many select lipid species have gained recognition for being crucial signaling molecules. For one example, the sphingolipid sphingosine-1-phosphate (S1P) is potently recognized by an array of S1P receptors which cause T cells to extravasate and chemotax toward sites of infection.

* Understanding the proteins that a given lipid interacts with can reveal great insight into the wide-ranging cellular signaling effects that a lipid can have.

## Why are functional modifications necessary in studying lipid interactions?

@@

@@

References to include: 
* [@schultzFlashClickMultifunctionalized2022a]
* [@schultzChemicalToolsLipid2023]
* [@farleyTrifunctionalSphinganineNew2024a]
* [@hoglingerTrifunctionalLipidProbes2017]

@@

#### Spectral counting

#### Isobaric mass tagging

\br

(Here's an example of what references would look like on the site: [@thomasTrifunctionalLipidDerivatives2025; @schultzFlashClickMultifunctionalized2022a; @schultzChemicalToolsLipid2023])

>>>>

---
title: "Multifunctional Lipid Probes Background"
author: "Gaelen Guzman"
format: 
  html:
    embed-resources: false
standalone: true
crossref: 
  ref-hyperlink: true
---

@@

## Introduction: Why Study Lipid Interactomes?

At its core, biology is a complex network of chemical interactions. Every biological process—from metabolism to signaling—depends on the physical interactions between molecules, including proteins, nucleic acids, and lipids. While proteins have long been the primary focus of interaction studies, lipids play an equally crucial yet often underappreciated role in cellular function.

### Lipid-Lipid Interactions and Membrane Properties
Lipids are primarily organized within membrane bilayers, where their interactions define critical membrane properties such as rigidity, fluidity, and resistance to detergents. By studying lipid-lipid interactomes, researchers can gain deeper insight into how these physical interactions influence membrane architecture and dynamics. Understanding these interactions is essential for unraveling fundamental processes like vesicle trafficking, cell signaling, and membrane compartmentalization.

### Lipids as Signaling Molecules
Beyond their structural roles, many lipids have been increasingly recognized as key signaling molecules. For example, the sphingolipid sphingosine-1-phosphate (S1P) plays a pivotal role in immune function. S1P binds to a specific family of receptors, triggering T cell migration (extravasation) and guiding their movement toward sites of infection. This highlights how lipid signaling can have far-reaching effects on cellular behavior and physiological responses.

### Lipid-Protein Interactions: Unlocking Cellular Communication

A single lipid species can interact with a diverse array of proteins, influencing everything from receptor activation to intracellular signaling cascades. Mapping these lipid-protein interactions provides a powerful approach to understanding how lipids exert their wide-ranging effects on cellular processes. By deciphering lipid interactomes, researchers can uncover novel regulatory mechanisms and potentially identify new therapeutic targets in diseases where lipid signaling is dysregulated.

## Synthetic Lipid Analogs: Powerful Tools for Lipid Biology
The study of lipid biology has been greatly advanced by synthetic lipid analogs. However, it may not be immediately clear why such analogs are so essential. Unlike proteins and nucleic acids, lipids present unique challenges that make traditional genetic tools less effective in studying their function.

### Why Lipids Require a Different Approach
Advances in genetic engineering—such as CRISPR and de novo gene synthesis—have revolutionized the study of nucleic acids and proteins. These techniques allow researchers to introduce precise modifications, such as:
* Altering enzyme activity by mutating active sites to change substrate specificity.
* Redirecting proteins by swapping subcellular localization sequences.
* Modifying protein interactions by swapping or removing binding domains.
* Visualizing proteins by fusing them with fluorescent tags (e.g., GFP).

Isolating protein complexes by adding affinity handles for purification.
In contrast, lipids pose distinct challenges that make such genetic manipulations impractical:
Lipids Are Synthesized Enzymatically
Unlike proteins, whose sequences are encoded in DNA, lipids are the product of enzymatic pathways. Their structures cannot be directly altered within a living system in the same way that gene sequences can be modified.
Early lipid research relied on gene knockouts and knock-ins to globally alter lipid composition, which provided broad insights into lipid classes but was less effective for studying individual lipid species.
Lipids Are Small and Structurally Sensitive
Compared to proteins, which consist of hundreds or thousands of atoms, lipids are much smaller (often composed of just a few dozen atoms).
Because of their size, even small modifications can dramatically alter their biophysical properties and biological function.
For example, unlike proteins, lipids cannot be tagged with large fluorescent proteins (e.g., GFP) or classical affinity tags (e.g., HA tags) without completely disrupting their activity.
Lipids Are Metabolically Dynamic
Lipids exist within a highly dynamic metabolic network, where they can be rapidly modified, cleaved, or converted into other lipid species.
Some lipids, like sphingosine, have extremely short half-lives—sometimes only minutes—making them difficult to track and study in real time.
The Role of Synthetic Lipid Analogs
To overcome these challenges, researchers have developed synthetic analogs of native lipids. These modified lipids allow for precise investigation of lipid function while preserving their essential biological properties. The key modifications used in this approach include:
Diazirine Rings – Enable covalent crosslinking to interacting proteins upon UV activation.
Terminal Alkynes – Allow for affinity enrichment using click chemistry.
Coumarin Photocages – Provide spatiotemporal control over lipid activity through light-triggered release.
These modifications stabilize protein-lipid interactions and enable the selective enrichment of lipid-bound proteins, all while maintaining tight spatial and temporal control. Each of these tools is described in further detail below.

@@

Lipid interactions are typically highly transient -- few opportunities for

References to include:

* [@schultzFlashClickMultifunctionalized2022a]
* [@schultzChemicalToolsLipid2023]
* [@farleyTrifunctionalSphinganineNew2024a]
* [@hoglingerTrifunctionalLipidProbes2017]

@@

@@

#### Label-Free Quantification
Label-free quantification (LFQ) is a widely used approach in mass spectrometry-based proteomics that estimates protein abundance without the need for chemical labeling. Instead, it relies on comparing the summed intensity of detected peptides across different samples analyzed under identical conditions.

##### How It Works
MS1 (Peptide Detection): Peptides are first isolated and detected based on their mass-to-charge ratio (m/z).
MS2 (Peptide Sequencing): Selected peptide ions are fragmented, and the resulting spectra are used to determine peptide sequences.
Quantification: The total signal intensity of all identified peptides corresponding to a given protein is summed, serving as a proxy for the protein’s absolute abundance.
Comparison Across Samples: When two or more samples are prepared and analyzed using the same LC-MS/MS protocol, their relative protein abundances can be compared.

##### Advantages of Label-Free Quantification
* *Cost-Effective & Simple Sample Preparation*
  * Unlike labeling methods (e.g., TMT), LFQ does not require expensive reagents or complex chemical modifications.
  * Sample preparation is straightforward, making LFQ accessible to a wide range of researchers.
* *Widely Used & Flexible*
  * LFQ is a common approach in proteomics, compatible with various experimental designs.
  * It allows for comparisons across a large number of samples without being limited by multiplexing constraints.

##### Challenges & Considerations
* *Potential Bias from Co-Elution:*
  * In complex samples, high-abundance peptides can overshadow low-abundance peptides, limiting the number of peptides that can be reliably isolated and quantified in the MS1 scan.
* *Run-to-Run Variability:*
  * Unlike TMT-based methods, where all samples are analyzed together in a single run, LFQ requires separate MS runs for each sample.
* Differences in instrument calibration, chromatography conditions, or sample preparation can introduce variability, potentially leading to artificial differences between samples if not carefully controlled.

##### Conclusion
Label-free quantification is a practical and widely used approach for proteomics research, offering a cost-effective and simple method for estimating protein abundance. However, careful experimental design and quality control are essential to minimize variability and ensure reliable quantification across different runs.

#### Quantitative Mass Spectrometry with Tandem Mass Tagging {#sec-TMT}

Tandem Mass Tag (TMT) labeling is a powerful technique in quantitative mass spectrometry (MS) that allows for the simultaneous analysis of multiple samples in a single MS run. This method enhances efficiency and enables precise relative quantification of proteins across different experimental conditions.

##### How It Works

* **Chemical Labeling**: Each peptide in a sample is chemically tagged with a unique TMT label, which consists of an identical reactive group that binds to peptides and a mass reporter region that differentiates samples.

* **Sample Pooling:** Once labeled, samples are combined into a single pooled sample group, reducing the need for separate MS runs.

* **Mass Spectrometry Analysis:** 
  * *MS1 (Peptide Isolation):* Peptides are first separated and detected based on their mass-to-charge ratio (m/z). 
  * *MS2 (Peptide Sequencing & Tag Fragmentation):* Peptides are fragmented, allowing for sequence identification. At the same time, the mass tags also fragment, producing unique reporter ions. 
  * *Quantification:* The intensity of these reporter ions reveals the relative abundance of each peptide across the original samples.

##### Advantages of TMT Labeling
* Increased Throughput & Efficiency
TMT allows for the multiplexing of many samples (e.g., 16 or more), reducing the number of separate MS runs needed.
Instead of running 16 individual samples separately, they can be pooled and analyzed together, increasing experimental efficiency.
* Deeper Proteome Coverage
The pooled sample can be further fractionated (e.g., using cation exchange chromatography), which helps separate high-abundance and low-abundance peptides before MS analysis.
This results in greater depth of peptide detection compared to running multiple low-depth, individual label-free MS analyses.
* Improved Quantification Accuracy
Since all samples are analyzed together in the same MS run, technical variability between runs is minimized, leading to more accurate relative quantification.

##### Challenges & Considerations
* Higher Cost
TMT reagents and the need for higher-end mass spectrometers contribute to increased costs.
* Complex Sample Preparation
Additional steps are required for labeling, pooling, and fractionation, which increases the risk of sample handling variability.
* Normalization Requirement
Total protein input must be carefully controlled to ensure accurate quantification. Unequal sample amounts (e.g., when comparing conditions with highly different protein yields) may require artificial balancing, which can introduce biases.
* Demanding Instrumentation & Expertise
The technique requires high-resolution mass spectrometers with precise scanning capabilities, often necessitating outsourcing to specialized facilities.

##### Conclusion
TMT-based quantitative mass spectrometry is a valuable tool for large-scale proteomics studies, providing high-throughput, precise, and deep proteome coverage. While it requires careful sample preparation and advanced instrumentation, its ability to multiplex multiple samples in a single run makes it an attractive option for researchers aiming for high-confidence protein quantification.
